---
title: "Reproducible Research Assignment 1"
author: "Sudhir"
date: "April 25, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)

```

## Loading the data


```{r readData}
setwd("C:/Users/Sugrutha/Documents/Appa/RProjects/Repoducable Research")
df <- read.csv("activity.csv", colClasses = c("numeric","Date","integer"))
str(df)

```

## What is mean total number of steps taken per day?

1] Calculate the total number of steps taken per day
2] Make a histogram of the total number of steps taken each day

```{r totalSteps}
dSteps <-  aggregate(list(Steps = df$steps), by=list(date=df$date), FUN=sum, na.rm=TRUE)
head(dSteps)

hist(dSteps$Steps, main="Histogram of Steps", col="red", xlab="Total steps per day")
 
```


3] Calculate and report the mean and median of the total number of steps taken per day

```{r meanMed}
sMean  <- mean(dSteps$Steps);
sMean;
sMedian <- median(dSteps$Steps);
sMedian;

```

## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r meanStepsPerIntval}
meanStepsPerIntval <- aggregate(df$steps ~ interval, df, FUN = "mean", na.rm = TRUE)

plot(meanStepsPerIntval, type="l", xlab="5 Minutes interval", ylab="Average number of steps")

```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r maxNumOfSteps}
maxNumOfSteps <- max(meanStepsPerIntval$`df$steps`)
maxNumOfSteps
meanStepsPerIntval[meanStepsPerIntval$`df$steps` == maxNumOfSteps,]

```

## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r SumNA}
sum(is.na(df$steps))

```

Devise a strategy for filling in all of the missing values in the dataset
(using the mean for that 5-minute interval)
 
 Add a new colunn to original dataset "FullSteps"

```{r FullSteps}
df$FullSteps <- df$steps
head(df)
```


loop thru all rows and replace NA with

```{r loopFillNA}

for (i in 1:nrow(df) ) {
  ## if steps is NA replace step value of 5-minute interval mean
  if (is.na(df$steps[i])) {
      intvalIndex <- which(meanStepsPerIntval$interval == df$interval[i]);
      df$FullSteps[i] <- meanStepsPerIntval$`df$steps`[intvalIndex];
  }
}
```


Create a new dataset with all he missing data filled in.

```{r newDf}

newDf <- df[, c(2,3, 4) ]
head(newDf)

```

Make a histogram of the total number of steps taken each day 

```{r histTotSteps}

TotalStepsTakenPDay <- aggregate( FullSteps ~ date, newDf, FUN = "sum" )
hist(TotalStepsTakenPDay$FullSteps, col="red",xlab="Steps", main="Total steps taken per day")

```

Calculate and report the mean and median total number of steps taken per day.

```{r meanMedTotalSteps}
eMean <- mean(TotalStepsTakenPDay$FullSteps);
eMean;

eMedian <- median(TotalStepsTakenPDay$FullSteps);
eMedian;
#head(newDf)

```
Do these values differ from the estimates from the first part of the assignment? 
What is the impact of imputing missing data on the estimates of the total daily number of steps?
Yes, they do differ.

```{r impactOfFillNA}
impactOnMean <- (eMean/sMean)*100 - 100;
impactOnMedian <- (eMedian/sMedian)*100 - 100;

```
Mean has increased by `r impactOnMean` percent, and Median has increased by `r impactOnMedian` percentage.

## Are there differences in activity patterns between weekdays and weekends?

Adding a new column with day of the week

```{r addDay}
newDf$day <- weekdays(newDf$date)
head(newDf)

```

Adding a new column to indicate if weekend or weekday

```{r addFactorWeek}
factWkDay <- c("Saturday","Sunday")
newDf$week <- ifelse ( newDf$day %in%  factWkDay, "Weekend", "Weekday" )
newDf$week <- as.factor(newDf$week )
head(newDf)

meanStepsPerIntvlByWeek <- aggregate(newDf$FullSteps, by = list(newDf$week, newDf$interval), mean )
head(meanStepsPerIntvlByWeek)
#Set column names
names(meanStepsPerIntvlByWeek) <- c("week","interval","steps")
head(meanStepsPerIntvlByWeek)
```

Plot the average steps taken on Week days and on Weekends separately
```{r xyplot}
library(lattice)

xyplot(meanStepsPerIntvlByWeek$steps ~ meanStepsPerIntvlByWeek$interval| meanStepsPerIntvlByWeek$week, layout=c(1,2) ,  type="l", xlab=" 5 minute interval", ylab="Average steps taken" )


```